Closing the empirical cycle

### 2023-02-01 action_letter.Rmd
update 2023-02-01 10:12:21

---
title       : "Using machine learning to identify early predictors of adolescent emotion regulation development"
authors     : "masked"
journal     : "Journal of Research on Adolescence"
editor      : "Raval"
manuscript  : "2022-182"
class       : "draft"
bibliography      : ["manuscript/r-references.bib", "manuscript/references.bib", "manuscript/Veni.bib"]
output      : papaja::revision_letter_pdf
---

```{r, message=FALSE, warning=FALSE}
library(revise)
man <- read_manuscript("manuscript/manuscript.Rmd")
letter <- read_manuscript("action_letter.Rmd")
mask = TRUE
```

@@

`r get_revision("priorlit", letter, quote = F)`

@@

`r get_revision(manuscript = man, "globalnorth")`

@@

`r get_revision(manuscript = man, "velez1")`

@@

`r get_revision(manuscript = man, "velezagosto")`

@@

`r get_revision(manuscript = man, "waves")`

@@

`r get_revision(manuscript = man, "dutch")`

@@

`r get_revision(manuscript = letter, "extractpreds", quote = F)`

@@

`r get_revision(manuscript = man, "parenting")`

@@

`r get_revision(manuscript = man, "parentrelated")`
</span>

@@

`r get_revision(manuscript = man, "overlap")`

@@

`r get_revision(manuscript = man, "cfi")`

@@

`r get_revision(manuscript = man, "missing")`
</span>

@@

`r get_revision(manuscript = man, "homogeneity")`

@@

`r get_revision(manuscript = man, "homogeneity2")`

@@

`r get_revision(manuscript = man, "stepiii")`

@@

`r get_revision(manuscript = man, "fullygrown")`

@@

`r get_revision(manuscript = man, "semforests")`

@@

`r get_revision(manuscript = man, "predictions")`

@@

`r get_revision(manuscript = man, "nonlinearexample")`

@@

`r get_revision(manuscript = man, "limitations")`

@@

`r get_revision(manuscript = man, "posthoc")`

@@

`r get_revision(manuscript = man, "errata")`

@@

`r get_revision(manuscript = man, "includepersonality")`

@@

Second, with regard to the Reviewer's statement that "the third-most important predictor being a positive
parenting behavior". Balanced relatedness is not a parenting practice (i.e., a specific action parents engage in as part of child rearing). Rather, it is an aspect of the (adolescent-perceived) emotional family climate. `r get_revision(manuscript = letter, "parents", quote = F)`

@@

`r get_revision(manuscript = man, "causality")`

@@

`r get_revision(manuscript = man, "singlefactor")`

@@

`r get_revision(manuscript = man, "motherfather")`

@@

`r get_revision(manuscript = man, "abstract")`

@@

`r get_revision(manuscript = man, "abundantlit")`
</span>

@@

`r get_revision(manuscript = man, "entropy")`

@@

`r get_revision(manuscript = man, "percentiles")`

@@

`r get_revision(manuscript = letter, "parents", quote = F)`

@@

`r get_revision(manuscript = man, "highrankingparent")`

@@

`r get_revision(manuscript = man, "riskassessment")`

@@

`r get_revision(manuscript = man, "culture")`

>>>>

---
title       : "Using machine learning to identify early predictors of adolescent emotion regulation development"
authors     : "masked"
journal     : "Journal of Research on Adolescence"
editor      : "Raval"
manuscript  : "2022-182"
class       : "draft"
mask: no
bibliography      : ["manuscript/r-references.bib", "manuscript/references.bib", "manuscript/Veni.bib"]
output      : papaja::revision_letter_pdf
---

```{r, message=FALSE, warning=FALSE}
library(revise)
test_env <- new.env()
read_manuscript("c:/git_repositories/veni_forest/manuscript/manuscript.Rmd", envir = test_env)
read_manuscript("action_letter.Rmd", envir = test_env)
mask = TRUE
```

@@

`r get_revision("priorlit", quote = F)`

@@

`r get_revision( "globalnorth")`

@@

`r get_revision( "velez1")`

@@

`r get_revision( "velezagosto")`

@@

`r get_revision( "waves")`

@@

`r get_revision( "dutch")`

@@

`r get_revision("extractpreds", quote = F)`

@@

`r get_revision( "parenting")`

@@

`r get_revision( "parentrelated")`
</span>

@@

`r get_revision( "overlap")`

@@

`r get_revision( "cfi")`

@@

`r get_revision( "missing")`
</span>

@@

`r get_revision( "homogeneity")`

@@

`r get_revision( "homogeneity2")`

@@

`r get_revision( "stepiii")`

@@

`r get_revision( "fullygrown")`

@@

`r get_revision( "semforests")`

@@

`r get_revision( "predictions")`

@@

`r get_revision( "nonlinearexample")`

@@

`r get_revision( "limitations")`

@@

`r get_revision( "posthoc")`

@@

`r get_revision( "errata")`

@@

`r get_revision( "includepersonality")`

@@

Second, with regard to the Reviewer's statement that "the third-most important predictor being a positive
parenting behavior". Balanced relatedness is not a parenting practice (i.e., a specific action parents engage in as part of child rearing). Rather, it is an aspect of the (adolescent-perceived) emotional family climate. `r get_revision("parents", quote = F)`

@@

`r get_revision( "causality")`

@@

`r get_revision( "singlefactor")`

@@

`r get_revision( "motherfather")`

@@

`r get_revision( "abstract")`

@@

`r get_revision( "abundantlit")`
</span>

@@

`r get_revision( "entropy")`

@@

`r get_revision( "percentiles")`

@@

`r get_revision("parents", quote = F)`

@@

`r get_revision( "highrankingparent")`

@@

`r get_revision( "riskassessment")`

@@

`r get_revision( "culture")`

### 2023-02-01 manuscript/manuscript.Rmd
update 2023-02-01 10:12:21

The first step towards theory formation is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020].
Phenomena considered relevant in a field of research are often summarized using literature reviews.
A recent review of 6305 publications on adolescents’ emotion regulation used text mining to identify relevant phenomena, including potential risk factors [@vanlissaMappingPhenomenaRelevant2021].
Narrative reviews are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008].
Text mining systematic reviews are more comprehensive and derive insights from the literature through a transparent and somewhat objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021].
This review thus offers a relatively comprehensive overview of potential predictors of emotion regulation development.
The present study used it as a starting point for an empirical investigation of the relative importance and direction of effect of these predictors.

@@

```{r}
desc_num <- read.csv("../desc_num.csv", stringsAsFactors = FALSE)
desc_cat <- read.csv("../desc_cat.csv", stringsAsFactors = FALSE)
```
<span id = "dutch">Participants were 497 Dutch adolescents and their parents enrolled in the ongoing longitudinal study `r ifelse(mask, "RADAR [@branjeResearchAdolescentDevelopment2018]", "**(masked reference)**")`, which 
received ethical approval from the ethical committee of the University Medical Centre `r ifelse(mask, "Utrecht", "**masked**")`.
The `r ifelse(mask, "RADAR", "**masked**")` sample was recruited from over 200 randomly selected elementary schools in the province of `r ifelse(mask, "Utrecht", "**masked**")`, and four main cities in `r ifelse(mask, "The Netherlands", "**masked**")`.</span>
Of 1,544 eligible families, 497 produced informed consent and were included in the study.
[From 2006 to 2011, trained interviewers completed six annual measurement waves by collecting questionnaire data during home visits.]{#waves}
All independent variables were assessed in the first measurement wave.
[Sample demographic characteristics are reported in Tables \@ref(tab:tabcat) and \@ref(tab:tabnum).]{#demographics}
The dependent variable was measured from the second to the sixth wave.
Families received €100 financial compensation for their participation in annual measurements.

@@

From wave 2-6, adolescents completed the Difficulties in Emotion Regulation Scale, which was reverse coded to represent emotion regulation [DERS, @gratzMultidimensionalAssessmentEmotion2004].
Based on prior psychometric analyses [see @vanlissaRoleFathersMothers2019],
we used 24 items of this 32-item scale
(e.g., "When I’m upset, I have difficulty thinking
about anything else").
Responses ranged
from 1 (Almost never) to 5 (Almost always).
Score reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
We used latent variable scores (instead of mean scores) for further analyses.
[To this end, we estimated a CFA with all DERS items loading on one latent variable per measurement wave.]{#singlefactor}
Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the sample average trajectory.
As measurement invariance must be imposed in order to compare latent scores across waves,
we compared a model with configural invariance ($BIC = 141247.46$)
to a model that imposed metric invariance ($BIC = 140913.27$).
The lower BIC indicates support for metric invariance.
The RMSEA and SRMR of the metric invariance model indicated acceptable fit ($\text{RMSEA} = .05$,  $\text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
[Note that the RMSEA has been found to be more appropriate for longitudinal data, possibly because the null-model on which the CFI is based is less appropriate for repeated measures [@leiteDetectingGrowthShape2011]]{#cfi}.
As score reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
we proceeded with individual factor scores for further analysis.

@@

The SEM-forests algorithm requires complete data and cannot accommodate multiple imputed data.
Consequently, we used single imputation using the missForest algorithm,
a random forests based approach with comparable performance to multiple imputation and FIML estimation [@stekhovenMissForestNonparametricMissing2012].
This method is valid, regardless of whether missingness is random or contingent on observed data (MAR or MCAR).
It does not make any distributional assumptions and does not extrapolate,
meaning it returns realistic values within the range of the observed data.
The proportion of missingness ranged from $`r report(min((495-desc_cat[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for categorical data (see Table \@ref(tab:tabcat)),
and from $`r report(min((495-desc_num[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for continuous data (see Table \@ref(tab:tabnum)).
[We compared participants with missing- versus complete data on demographic characteristics (sex, age, socio-economic status, religion and religiosity) and scores of emotion regulation in all waves, and found no significant differences, $p = [0.06, 0.48]$.]{#missing}
<!-- Original publication had up to 30% missing, Wolputte, E. V., & Blockeel, H. (2020, October). Missing value imputation with MERCS: a faster alternative to MissForest. In International Conference on Discovery Science (pp. 502-516). Springer, Cham. looked at higher numbers. According to Mplus: minimum 10% -->

>>>>

The first step towards theory formation is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020].
Phenomena considered relevant in a field of research are often summarized using literature reviews.
A recent review of 6305 publications on adolescents’ emotion regulation used text mining to identify relevant phenomena, including potential risk factors [@vanlissaMappingPhenomenaRelevant2021].
Narrative reviews are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008].
Text mining systematic reviews are more comprehensive and derive insights from the literature through a transparent and somewhat objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021].
The present study used this prior text mining review by Van Lissa as a starting point for an empirical investigation of the relative importance and direction of effect of these predictors.
It provides a relatively comprehensive overview of potential predictors of emotion regulation development.

@@

```{r}
desc_num <- read.csv("../desc_num.csv", stringsAsFactors = FALSE)
desc_cat <- read.csv("../desc_cat.csv", stringsAsFactors = FALSE)
```

<span id = "dutch">Participants were 497 Dutch adolescents and their parents enrolled in the ongoing longitudinal RADAR study  [@branjeResearchAdolescentDevelopment2018], which 
received ethical approval from the ethical committee of the University Medical Centre Utrecht.
The RADAR sample was recruited from over 200 randomly selected elementary schools in the province of Utrecht, and four main cities in The Netherlands.</span>
Of 1,544 eligible families, 497 produced informed consent and were included in the study.
[From 2006 to 2011, trained interviewers completed six annual measurement waves by collecting questionnaire data during home visits.]{#waves}
All independent variables were assessed in the first measurement wave.
[Sample demographic characteristics are reported in Tables \@ref(tab:tabcat) and \@ref(tab:tabnum).]{#demographics}
The dependent variable was measured from the second to the sixth wave.
Families received €100 financial compensation for their participation in annual measurements.

@@

From wave 2-6, adolescents completed the Difficulties in Emotion Regulation Scale, which was reverse coded to represent emotion regulation [DERS, @gratzMultidimensionalAssessmentEmotion2004].
Based on prior psychometric analyses [see @vanlissaRoleFathersMothers2019],
we used 24 items of this 32-item scale
(e.g., "When I’m upset, I have difficulty thinking
about anything else").
Responses ranged
from 1 (Almost never) to 5 (Almost always).
Score reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
We used latent variable scores (instead of mean scores) for further analyses.
[To this end, we estimated a CFA with all DERS items loading on one latent variable per measurement wave.]{#singlefactor}
Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the sample average trajectory.
As measurement invariance must be imposed in order to compare latent scores across waves,
we compared a model with configural invariance ($BIC = 141247.46$)
to a model that imposed metric invariance ($BIC = 140913.27$).
The lower BIC indicates support for metric invariance.
The RMSEA and SRMR of the metric invariance model indicated acceptable fit ($\text{RMSEA} = .047$,  $\text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
[Note that the RMSEA has been found to be more appropriate for longitudinal data, possibly because the null-model on which the CFI is based is less appropriate for repeated measures [@leiteDetectingGrowthShape2011]]{#cfi}.
As score reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
we proceeded with individual factor scores for further analysis.

@@

The SEM-forests algorithm requires complete data and cannot accommodate multiple imputed data.
Consequently, we used single imputation using the missForest algorithm,
a random forests based approach with comparable performance to multiple imputation and FIML estimation [@stekhovenMissForestNonparametricMissing2012].
This method is valid, regardless of whether missingness is random or contingent on observed data (MCAR or MAR).
It does not make any distributional assumptions and does not extrapolate,
meaning it returns realistic values within the range of the observed data.
The proportion of missingness ranged from $`r report(min((495-desc_cat[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for categorical data (see Table \@ref(tab:tabcat)),
and from $`r report(min((495-desc_num[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for continuous data (see Table \@ref(tab:tabnum)).
[We compared participants with missing- versus complete data on demographic characteristics (sex, age, socio-economic status, religion and religiosity) and scores of emotion regulation in all waves, and found no significant differences, $p = [0.06, 0.48]$.]{#missing}
<!-- Original publication had up to 30% missing, Wolputte, E. V., & Blockeel, H. (2020, October). Missing value imputation with MERCS: a faster alternative to MissForest. In International Conference on Discovery Science (pp. 502-516). Springer, Cham. looked at higher numbers. According to Mplus: minimum 10% -->
